---
title: "Ch9_2ndEdition"
output: html_document
---

```{r}
library(rethinking)
```

# Chapter 9 - Markov Chain Monte Carlo

Used for estimation of posterior distributions
* Produce samples from the joint posterior w/out maximizing anything
* Sample directly from teh posterior 


## 9.1 Good King Markov & His Island Kingdom
(Metropolis algorthim)
* Circle of islands 1-10
* Can only travel to adjacent islands
* wants to visit islands in proportion to population

(1) Each week, he decides to stay or move. 
(2) Coin flip determines proposal island (left or right)
(3) Count out # of seashells for population of both islands
(4) If new island > current, he moves. If new island < current, p(move) = #new/#current

Procedure guarantees that w/enough sampling he will visit island w/ the correct proportion


## 9.2 Metropolis Algorithm 
*used to draw samples from a complex target distribution.
To convert from the previous island analogy: 
*"islands" -> our objective parameter values
*"population sizes" -> posterior probabilities at each parameter value
*"weeks" -> samples taken from the joint posterior of the parameters

### 9.2.1 
Gibbs > Metropolis, but there are better/other options:
  * Metropolis-Hastings sampling: Bias the coin toss to lead him to higher posterior probabilities
  * Gibbs sampling: "Adaptive proposals" - distribution of proposed parameters adjust intelligently (using conjugate pairs)
  
### 9.2.2 - High Dimensional Sadness
:( Gibbs Sampling:
  *What if I don't want to use conjugate pairs?
  *1000 - 10000 parameters hurts Gibbs.
  *Regions of high correlation hurt -> a narrow ridge of high probability
  
"Conentration of Measure" - prob. mass of high-dimension distance is far from mode of distribution. Wont get points near the mode, but in the probability mass.

(1) Need MCMC algorithms that focus on entire posterior
(2) Optimization models that look for mode (quap) are no use at all

It can be dangerous to summarize parameter distributions w/ mode. Need to be hunting for probability density outside of that narrow range.

## 9.3 Hamiltonian Monte Carlo

### 9.3.1 Extension of the MC Analogy
Valley w/ population proportional to elevation (most people live in valley)

* Vehicle picks a random direction w/ random momentum
* As it goes uphill, it's forced to stop & turn around
* After fixed period of time, get out & look around
* Locations visited inversely proportional to elevation
* Low autocorrelation between locations visited
* Requires a continuous (land, parameters?) to do this 

### 9.3.2 Particles in Space
* HMCMC literally simulated by a frictionless partical in space
  - almost every proposal accepted (~95%) because it makes intelligent proposals
  - Energy in system should be "conserved" (reject these proposals)
  
toy example w/ X&Y values simulated from the normal

Functions: 
(1) computes log-probability of the data & parameters 
  * Top part of Bayes Eqn
  * Tells the "elevation" of any set of parameter values
(2) Needs the gradient
  * aka: the derivative of x & y
  
Settings:
(1) # of Leapfrog Steps:
  * each curve/path is divided into steps:
    - many steps -> long paths
    - few steps -> short paths
(2) Step Size:
  * How fine grained a simulation is
  
Allow the sampler to warmup first -> try to figure out what is a good step size. This is different from burn-in because we aren't actually sampling here.

No-U-Turn samples - adaptively set the number of leapfrog steps

MCMC needs 5 things to go:
(1) Function U that returns the negative log-probability of the data at the current position
(2) Function grad_U that returns the gradient
(3) step size epislon
(4) count of leapfrog steps (L)
(5) Starting position

### 9.3.3 Limitations:
*requires continuous parameters

## 9.4 Easy MCMC: ulam

Uses same helper functions (extract functions, extract prior, link, sim)
```{r}
library(rethinking)
data(rugged)
d <- rugged
d$log_gdp <- log(d$rgdppc_2000)
dd <- d[ complete.cases(d$rgdppc_2000) , ]
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
dd$rugged_std <- dd$rugged / max(dd$rugged)
dd$cid <- ifelse( dd$cont_africa==1 , 1 , 2 )
```

```{r}
m8.5 <- quap(
    alist(
        log_gdp_std ~ dnorm( mu , sigma ) ,
        mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) ,
        a[cid] ~ dnorm( 1 , 0.1 ) ,
        b[cid] ~ dnorm( 0 , 0.3 ) ,
        sigma ~ dexp( 1 )
),
    data=dd )
precis( m8.5 , depth=2 )
```


### 9.4.1 Preparation
Use same focumula as before, but do two additional things
(1) Preprocess all variable transformations
(2) Trim off to only used variables
  (2.5) put things into a list instead of a df

```{r}
dat_slim <- list(
    log_gpd_std = dd$log_gdp_std,
    rugged_std = dd$rugged_std,
    cid = as.integer( dd$cid )
)
str(dat_slim)
```

### 9.4.2 Sampling from the posterior
Looks just like before 
pecis()
Rhat should converge to 1.0
n_eff is a crude estimate of sample size

```{r}
m9.1 <- ulam(
    alist(
        log_gdp_std ~ dnorm( mu , sigma ) ,
        mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) ,
        a[cid] ~ dnorm( 1 , 0.1 ) ,
        b[cid] ~ dnorm( 0 , 0.3 ) ,
        sigma ~ dexp( 1 )
),
data=dat_slim , chains=1 )
```

```{r}
precis( m9.1 , depth=2 )
```
ulam doesn't seem to be working super well here

### 9.4.3 Sampling again, in parallel
can choose number of chains and cores
```{r}
m9.1 <- ulam(
    alist(
        log_gdp_std ~ dnorm( mu , sigma ) ,
        mu <- a[cid] + b[cid]*( rugged_std - 0.215 ) ,
        a[cid] ~ dnorm( 1 , 0.1 ) ,
        b[cid] ~ dnorm( 0 , 0.3 ) ,
        sigma ~ dexp( 0.1 )
),
data=dat_slim , chains=4 , cores=4 , iter=1000)
```
```{r}
show(m9.1)
```

```{r}
precis( m9.1 , 2 )
```

Something real not great is happening here...

### 9.4.4 Visualization
```{r}
pairs(m9.1)
```


### 9.4.5 Checking the chain
Look at the trace plot:
(1) Stationarity - path stays w/in high probability portion of the distribution
(2) Good Mixing - Rapidly zig-zags, explores the region well, not slowly drifting
(3) Convergence - Multiple chains stick in the same region

```{r}
traceplot(m9.1)
```


Trank (Trace Rank) Plot - Very similar, just ranked version of trace plot w/ multiple chains. Lots of intermingling of chains is a good sign.
```{r}
trankplot(m9.1, n_cols = 2)
```


## 9.5 Care and Feeding of Your Markov Chain

### 9.5.1 How many samples do you need?
(1) n_eff is what matters in the end
(2) depends on what you want to know:
  * to learn posterior means you don't need a lot
  * to learn shape of the posterior tail at 99th percentile, you need many more
  * n_eff = 200 for most regressions
  
### 9.5.2 How many chains do you need?
(1) Debugging a model - use a single chain
(2) Deciding if a chain is valid - use more than one (3-5 is what I've seen)
(3) Final run for inference - run 1 long chain

"Four short chains to check, one long chain for inference"
If rhat > 1.00, chain may not have converged

### 9.5.3 Taming a wild chain
* putting flat priors can cause problems - give it some suggestive prior

```{r}
y <- c(-1,1)
set.seed(11)
m9.2 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- alpha ,
        alpha ~ dnorm( 0 , 1000 ) ,
        sigma ~ dexp( 0.0001 )
),
data=list(y=y) , chains=2 )
```

```{r}
precis(m9.2)
```

```{r}
traceplot(m9.2)
```

add the weakly informed prior
```{r}
set.seed(11)
m9.3 <- ulam(
alist(        y ~ dnorm( mu , sigma ) ,
        mu <- alpha ,
        alpha ~ dnorm( 1 , 10 ) ,
        sigma ~ dexp( 1 )
),
    data=list(y=y) , chains=2 )
precis( m9.3 )
```

```{r}
traceplot(m9.3)
```


### 9.5.4 Non-identifiable parameters
* Commonly caused by highly correlated predictors
```{r}
set.seed(41)
y <- rnorm( 100 , mean=0 , sd=1 )
```

```{r}
m9.4 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- a1 + a2 ,
        a1 ~ dnorm( 0 , 1000 ),
        a2 ~ dnorm( 0 , 1000 ),
        sigma ~ dexp( 1 )
),
data=list(y=y) , chains=2 )
```

```{r}
precis( m9.4 )
```

Regularizing weak priors help us (but I wish I could get them to help above)
```{r}
m9.5 <- ulam(
    alist(
        y ~ dnorm( mu , sigma ) ,
        mu <- a1 + a2 ,
        a1 ~ dnorm( 0 , 10 ),
        a2 ~ dnorm( 0 , 10 ),
        sigma ~ dexp( 1 )
),
    data=list(y=y) , chains=2 )
precis( m9.5 )
```

```{r}
traceplot(m9.4)
traceplot(m9.5)
```

